Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
========================================================================================
   
**Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)

**July 17, 2014**

**Technical Report No. 3**

**Department of Biology**

**University of Vermont**

```{r setup, echo=FALSE, results='hide', message = FALSE}
# Global settings
options(stringsAsFactors=FALSE, digits = 3, scipen = 1)

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)
library(MASS)
library(xtable)
library(methods) # required by `knitcitations` when file is `knit` from command-line
library(reshape2)
library(RColorBrewer)

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
#biocLite("GOSemSim")
library(topGO) 
library(Rgraphviz)	
library(GOSemSim)

# load custom functions
source("scripts/RxNseq.R")

# knitr options
opts_chunk$set(cache=TRUE)

# directory to save results
resultsdir <- "results/"
```

## Summary ##
  
In this technical report, which accompanies the manuscript **Patterns of stress tolerance versus resistance in genome-wide expression data of parapatric ant species** (Stanton-Geddes et al., in press), we:

1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
2. identify thermally-responsive genes
3. evaluate differences in reaction norms of expression for each thermally-responsive gene between the two species
4. perform gene set enrichment analysis of thermally-responsive genes for the two species

This script is completely reproducible assuming that R, `knitr` and the other required libraries (listed within the source document) are installed on a standard linux system using the following:
    
    Rscript -e "library(knitr); knit('ApTranscriptome_TR.Rmd')"





## Data ##

The raw Illumina fastq files are available from the NCBI short read archive [link tbd]. The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.

## Sample description ##

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice Bacon DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*. For historical reasons, I refer to these species as colonies at times throughout this technical report.

For each colony, three ants were exposed to one of 12 temperature treatments, every 3.5C ranging from 0C to 38.5C, for one hour in glass tubes in a water bath. The ants were flash frozen and stored at -80C until RNA was extracted using a two step extraction; [RNAzol RT](http://www.mrcgene.com/rnazol.htm) (Molecular Research Center, Inc) followed by an [RNeasy Micro](http://www.qiagen.com/products/catalog/sample-technologies/rna-sample-technologies/total-rna/rneasy-micro-kit) column (Qiagen). Samples from each colony were pooled and sequenced in separate lanes on a 100bp paired-end run of an Illumina HiSeq at the University of Minnesota Genomics Center, yielding 20e6 and 16e6 reads for the A22 and Ar samples, respectively.



## Transcriptome assembly ##

The Illumina reads were filtered using the program [Trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) `r citep("10.1093/nar/gks540")` to remove Ilumina adapter sequences, trim bases with PHRED quality scores less than 15 in a 4 bp sliding window and remove final trimmed sequences with length less than 36 bp. The code used was 

~~~
java -jar trimmomatic-0.30.jar PE -threads 40 -phred33 -trimlog trimmomatic.log sample.R1.fastq sample.R2.fastq sample.R1.trimmed.paired.fastq sample.R1.trimmed.unpaired.fastq sample.R2.trimmed.paired.fastq sample.R2.trimmed.unpaired.fastq ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:36
~~~

where "sample" was replaced by each sample. This filtering yielded 339,845,787 properly paired reads and 16,860,885 unpaired reads. 

Properly paired and unpaired reads passing the Trimmomatic filter were combined and used in de novo transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`. Unpaired reads resulting from Trimmomatic were concatenated to the set of left (R1) reads per Trinity usage documentation. Assembly on a single 40 CPU machine with 1TB of memory required 20 hours of compute time.

~~~
Trinity.pl --seqType=fq --JM=90G —left=all.R1.trimmed.fastq —right=all.R2.trimmed.fastq --output=trinity --CPU=40 --inchworm_cpu=40 --bflyCPU=5
~~~

This assembly contained 100,381 unique components (roughly genes) in 126,172 total transcripts (Table 1). 

As we were assembling two divergent colonies into a single transcriptome, we suspected that this assembly would be susceptible to known problems of errors during assembly (e.g. chimeric transcripts that are fusions of two transcripts) and redundancy `r citep("10.1186/1471-2164-14-328")`. To account for this, we performed two post-assembly processing steps.

First, we ran the program [cap3](http://seq.cs.iastate.edu/) `r citep("10.1101/gr.9.9.868")` setting the maximum gap length and band expansion size to 50 `-f 50 -a 50`, no end clipping as the reads were already filtered `k 0`, requiring 90% identity for assembly, and a minimum overlap length of 100 bp `-o 100`. The percent identity threshold of 90% was chosen to liberally collapse orthologous contigs from the two colonies that may have been assembled separately. 

    cap3 Trinity.fasta -f 50 -a 50 -k 0 -p 90 -o 100 > Trinity_cap3.out

The output of `cap3` gives assembled "contigs" and unassembled "singlets" that were concatenated into a single file.

    # check the number of contigs clustered
    grep -c "Contig" Trinity.fasta.cap.contigs
    grep -c "comp" Trinity.fasta.cap.singlets
    # compare to contigs from Trinity output
    grep -c "comp" Trinity.fasta

    # Combine contigs and singlets from CAP3
    cat Trinity.fasta.cap.contigs Trinity.fasta.cap.singlets > Trinity_cap3.fasta

The output file "Trinity.fasta.cap.info" gives specific information on which contigs were collapsed.

Subsequent to running `cap3`, we ran [uclust](http://drive5.com/usearch/manual/uclust_algo.html) to cluster sequences completely contained within longer sequences, again specificing a 90% identity cutoff for clustering. 

    # sort
    uclust --sort Trinity_cap3.fasta --output Trinity_cap3_sorted.fasta
    # cluster by 90% similarity threshold
    uclust --input Trinity_cap3_sorted.fasta --uc Trinity_cap3_uclust.out --id 0.90
    # convert uclust to fasta format
    uclust --uc2fasta Trinity_cap3_uclust.out --input Trinity_cap3_uclust.fasta

These post-processing step removed `r round((126172-105536)/126172 * 100, 0)`% of the initial reads (Table 1).

```{r assemstats, results = "asis", echo=FALSE}
# make table
trinity <- c("126,172", "100,389,539", "358", "795", "16,201", "1,631")
reduced <- c("105,536", "62,648,997", "320", "593", "15,491", "895")

assemstats <- rbind(trinity, reduced)
colnames(assemstats) <- c("Total contigs", "Total length", "Median contig size", "Mean contig size", "N50 contig", "N50 Length")

pandoc.table(assemstats, style="rmarkdown", caption = "Statistics for Trinity and cap3+uclust reduced transcriptome assemblies")
```

To remove contigs that are likely contaminants from bacterial, archael, virus or human sources, we used the program [DeconSeq](http://deconseq.sourceforge.net/) r citep("10.1371/journal.pone.0017288"). We downloaded the bacteria, virus, archae and human [databases of contaminants](ftp://edwards.sdsu.edu:7009/deconseq/db), modified the `DeconSeqConfig.pm` file as described [here](http://www.vcru.wisc.edu/simonlab/bioinformatics/programs/install/deconseq.htm) to point to the databases, and ran DeconSeq specifiying 95% identity over 50% the length of contig

    deconseq.pl -c 50 -i 95 -f Trinity_cap3_uclust.fasta -d Trinity_cap3_uclust -dbs hsref,bast,vir,arch
	
This resulted in removing 5,675 contigs as contaminants, leaving 99,861 "clean" contigs. We spot-checked the contaminants by BLAST and confirmed that they matched bacteria, human or viral sources by greater than 95%. For expression quantification, we use the full assembly to ensure that "contaminant" reads are assigned to the contaminants. After quantification, these transcripts will then be removed from further analyses.

Running Trinity and subsequent programs is time and memory-intensive so the final assembly can be downloaded and used for all further analyses. In addition, the compressed archive contains the "clean" and "contaminant" sequences after filtering with DeconSeq.

~~~
# download filtered Trinity assembly, uncompress and move
wget http://johnstantongeddes.org/assets/files/Aphaenogaster_transcriptome.tar
# check md5sum
md5sum Aphaenogaster_transcriptome.tar
# fd6dbb0b3e88e1c797c9e74611b245b2

# uncompress and move
tar -xvf Aphaenogaster_transcriptome.tar
mkdir -p results/
mkdir -p results/trinity-full/
mv Trinity_cap3_uclust.fasta results/trinity-full/.
mv Trinity_cap3_uclust_clean.fasta results/trinity-full/.
~~~

To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) r citep("doi:10.1186/1471-2105-12-385"). I ... 

~~~
KRONA code
~~~

The interactive visualization is available [here]().


## Transcriptome annotation ##

Annotation was performed by uploading the reduced assembly "Trinity_cap3_uclust.fasta" to the web-based annotation program [FastAnnotator](http://fastannotator.cgu.edu.tw/index.php).

Results are available as job ID [13894410176993](http://fastannotator.cgu.edu.tw/job.php?jobid=13894410176993#page=basicinfo).

This annotation file can be read directly to R:

```{r download, echo = TRUE}
# URL for annotation file
annotation.URL <- getURL("http://johnstantongeddes.org/assets/files/ApTranscriptome_AnnotationTable_20140113.txt")
# load 
annotation.file <- read.csv(textConnection(annotation.URL), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
str(annotation.file)

# Convert to data.table
annotation.table <- data.table(annotation.file)
str(annotation.table)
```

Note that I used the reduced assembly, prior to cleaning out contaminants. As I use the "clean" assembly for read mapping and identification of responsive genes, the contaminants are dis-regarded in downstream analyses. 


## Identification of thermally-responsive genes ##

### Quantify gene expression ###

I quantified gene expression using [sailfish](http://www.cs.cmu.edu/~ckingsf/software/sailfish/index.html). To run this program, first make sure that PATHs to the software libraries are set up correctly as described on the sailfish website. 
                                                 
An index of the assembly is built with the command:                                                 

~~~
sailfish index -t results/trinity-full/Trinity_cap3_uclust.fasta -o results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4
~~~

Once this is done, expression is quantified for the Trimmomatic filtered reads from each species-treatment sample separately. Note that for each sample, there are four filtered read files:

- paired.left.fastq
- paired.right.fastq
- unpaired.left.fastq
- unpaired.right.fastq



~~~
# make a directory for the expression values
mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust
# change to "trinity-full" directory
cd results/trinity-full
# for each sample, run the following command
sailfish quant -i sailfish-index-Trinity-cap3-uclust -o sailfish-expression-Trinity-cap3-uclust/A22-0 -l "T=SE:S=U" -r A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq A22-0_ATCACG.paired.left.fastq A22-0_ATCACG.paired.right.fastq -p 4
~~~
                                           
While it is possible to separately specify the paired-end and orphaned single-end reads in Sailfish v0.6.3, the results are exactly the same as if they are all entered as SE.	

These files are downloaded for convenience:

~~~
wget ...
mkdir -p /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
mv .. /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
tar -zxvf ...
~~~

For each sample, there is a directory containting a file *quant_bias_corrected.sf*. This file has the following columns, following a number of header lines:

1. Transcript ID
2. Transcript Length
3. Transcripts per Million (TPM): computed as described in `r citep("10.1093/bioinformatics/btp692")`, and is meant as an estimate of the number of transcripts, per million observed transcripts, originating from each isoform.
4. Reads Per Kilobase per Million mapped reads (RPKM): classic measure of relative transcript abundance, and is an estimate of the number of reads per kilobase of transcript (per million mapped reads) originating from each transcript.

The TPM column for each sample was extracted and combined into a matrix for each species.

**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.** 


```{r load_expression_data, eval=TRUE, echo=FALSE, results='hide'}
# sailfish expression directory
sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"

samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")
    
# read in each file using loop
for (j in 1:length(samples)) {
    samp <- samples[j]
    trtval <- as.numeric(str_split_fixed(samp, "-", 2)[2])
    outpre <- gsub("-", "_", samp)
    outname <- paste(outpre, "_quant", sep="")
    read.sailfish.quant(filein=paste(sfexpressionroot, samp, "_quant/quant_bias_corrected.sf", sep=""), outname=outname, samp = samp, trtval = trtval)
}

# combine into long format data.table
A22.TPM <- data.table(rbind(A22_0_quant, A22_3_quant, A22_7_quant, A22_10_quant, A22_14_quant, A22_17_quant, A22_21_quant, A22_24_quant, A22_28_quant, A22_31_quant, A22_35_quant, A22_38_quant))
A22.TPM[,colony:="A22"]

Ar.TPM <- data.table(rbind(Ar_0_quant, Ar_3_quant, Ar_7_quant, Ar_10_quant, Ar_14_quant, Ar_17_quant, Ar_21_quant, Ar_24_quant, Ar_28_quant, Ar_31_quant, Ar_35_quant, Ar_38_quant))
Ar.TPM[,colony:="Ar"]

# combine colonies into single data.frame
TPM.dt <- rbind(A22.TPM, Ar.TPM)
TPM.dt$colony <- as.factor(TPM.dt$colony)
str(TPM.dt)

# drop temperature 7 
setkey(TPM.dt, val)
TPM.dt.sub <- TPM.dt[val != 7] 

length(unique(TPM.dt.sub$Transcript))

# set "trt" to true values - truncated in file names for convenience
TPM.dt.sub[val==3,val:=3.5]
TPM.dt.sub[val==10,val:=10.5]
TPM.dt.sub[val==17,val:=17.5]
TPM.dt.sub[val==24,val:=24.5]
TPM.dt.sub[val==31,val:=31.5]
TPM.dt.sub[val==38,val:=38.5]
str(TPM.dt.sub)
unique(TPM.dt.sub$val)
```

Note that expression levels at each temperature treatment are highly correlated between the two colonies.

```{r exp_correlations, echo=FALSE, results='hide'}
cortable <- ddply(TPM.dt.sub, .(val), summarize, cor(TPM[which(colony=="A22")], TPM[which(colony=="Ar")]))
```

```{r exp_correlations_table, echo=FALSE, results='asis'}
colnames(cortable) <- c("Temperature", "r")
pandoc.table(cortable, style="rmarkdown", caption = "Correlations between species for gene expression at temperature treatment", round=2)
```


### Remove *contaminant* transcripts

With gene expression quantified using all reads, I now remove the transcripts identified as contaminants by DeconSeq.

```{r remove_contaminants}
# extract transcript names from fasta file rather than loading whole file
system("grep '^>' results/trinity-full/Trinity_cap3_uclust_cont.fa | cut -f 1 -d ' ' | sed -r 's/^.{1}//' > results/trinity-full/cont.list")

# read file
cont.list <- read.table("results/trinity-full/cont.list")
str(cont.list)

# remove from TPM.dt.sub
setkey(TPM.dt.sub, Transcript)
TPM.dt.sub <- TPM.dt.sub[!cont.list]
str(TPM.dt.sub)
length(unique(TPM.dt.sub$Transcript))

# remove from annotation.table
setkey(annotation.table, Sequence.Name)
annotation.table <- annotation.table[!cont.list]
str(annotation.table)
``` 

### Regression-model to identify thermally-responsive genes

To identify transcripts (roughly equivalent to genes) that show thermal responsiveness, I fit the following linear model to each transcript:

$$ log(TPM + 1) = \beta_0 + \beta_1(species) + \beta_2(temp) + \beta_3(temp^2) + \beta_4(species * temp) + \beta_5(species * temp^2) + \epsilon $$

where TPM is transcripts per million. 


(1) Identify transcripts with overall significant model fit. Adjust *P* values for multiple testing using FDR and retain transcripts with FDR < 0.05. Use log-transformed response to account for outliers.


```{r modp}
# define model for RxN function
model <-  "log(TPM+1) ~ colony + val + I(val^2) + colony:val + colony:I(val^2)"

# calculate overall P value and R^2 for each transcript
RxNpval <- ddply(TPM.dt.sub, .(Transcript), .inform="TRUE", modpFunc)
```

Of the `r nrow(RxNpval)` transcripts, `r length(which(RxNpval$pval < 0.05))` have models with P < 0.05.

Many of these are likely false positives, so I adjust P-values using false discovery rate (FDR). Only those transcripts with less than 5% FDR are retained as significant. 

```{r padjust}
RxNpval$padj <- p.adjust(RxNpval$pval, method = "fdr")
# Plot FDR values against initial pvalues
par(mfrow = c(2,1))
hist(RxNpval$pval)
hist(RxNpval$padj)

# subset to significant transcripts
signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]

# extract significant transcripts
sig.TPM.dt.sub <- TPM.dt.sub[signif.transcripts$Transcript]
```

At the 5% FDR significance threshold, there are `r nrow(signif.transcripts)` transcripts with an overall significant model.


(2) Fit linear model to overall significant transcripts; perform stepAIC to retain only significant terms, and save `lm` output to list

```{r RxNlmAIC, echo=TRUE, eval=TRUE, results='hide'}
# perform model selection for responsive transcripts
# need to use `try` to avoid stopping on error for AIC at Infinity
RxNlmAIC <- try(dlply(sig.TPM.dt.sub, .(Transcript), lmFunc))
```


### Grouping of thermally-responsive transcripts

The set of transcripts with significant expression patterns include those with expression that differs by species, temperature and the interaction of species and temperature. In this section, I am specifically interested in the thermally-responsive transcripts (temperature and species x temperature) so I subset the significant transcripts to examine these. 

```{r responsive_transcripts}
interaction.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
other.lms <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.lms))]
temperature.lms <- other.lms[which(Map(grepFunc, other.lms, term = "val") == TRUE)]
colony.lms <- other.lms[setdiff(names(other.lms), names(temperature.lms))]
responsive.lms <- c(temperature.lms, interaction.lms)
rm(other.lms)
```

```{r table_results, echo=FALSE, results='asis'}
clist <- c('Total', 'Colony', 'Temperature', 'Temperature:Colony')
siglist <- c(length(RxNlmAIC),
             length(colony.lms),
             length(temperature.lms),
             length(interaction.lms)
             )

sigtable <- data.frame(Coefficient = clist, "Number significant" = siglist)

pandoc.table(sigtable, style="rmarkdown", caption = "Number of transcripts with expression that depends on species, temperature or their interaction at 5% FDR  out of 98,186 total transcripts.")
```


### Thermal-response functional types ###

The previous section identified the transcripts with thermally-responsive expression. In this section, I determine the shape of the expression response to temperature for each transcript. Catego
ries of expression response are:

* High - increase expression with temperature
* Low - decrease expression with temperature
* Intermediate - maximum expression at intermediate temperatures (14 - 28C)
* Bimodal - expressed greater than one standard deviation of expression at both low and high temperatures

I do this first for the thermally-responsive transcripts where there is no interaction with species. For the transcripts where thermal-responsive expression depends on species, I determine the functional type of the expression response separately for each species. 


```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for responsive transcripts
interaction.response.type <- ldply(interaction.lms, .progress = "none", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.lms))

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.lms, .progress = "none", .inform = TRUE, RxNtype)
stopifnot(nrow(temperature.response.type) == length(temperature.lms))

# merge results
Ap.response.type <- rbind(interaction.response.type, temperature.response.type)
colnames(Ap.response.type)[which(colnames(Ap.response.type) == ".id")] <- "Transcript"
str(Ap.response.type)

# save results to file
write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), Ap.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```

Next, I compare the number of thermally-responsive in each response category between the two colonies. 

```{r temperature_response}
A22.type.table <- table(Ap.response.type[ , 'A22.type'])
Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])

# table
(Ap.type.table <- rbind(A22.type.table, Ar.type.table))

# Pearson Chi-square test
chi1 <- chisq.test(Ap.type.table)
chi1

# Reorganize for plotting to show overlap among categories between colonies
type.table <- table(Ap.response.type[ , 'Ar.type'], Ap.response.type[ , 'A22.type'])
# reorder
tt2 <- type.table[c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", "Intermediate", "High", "Bimodal", "NotResp")]

# plot
png("mosaic_plot.png")
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4], "A. carolinensis", xpd = TRUE, font = 3)
legend(1.05, .75, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
dev.off()
```

The number of thermally-responsive in each response category differs between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies. For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts. For *AcNC*, transcripts expressed at *Intermediate* temperatures are next most common, followed by *High* and *Bimodal*. 

Interestingly, nearly half of the *High* genes in *AcNC* are *Low* in *ApVT*, and vice versa. In contrast, most of the *Low* genes in one species are also *Low* in the other species.


```{r temperature_response_table, echo=FALSE, results='asis'}
# add row and column totals
tt3 <- rbind(tt2, colSums(tt2))
tt3 <- cbind(tt3, rowSums(tt3))
colnames(tt3)[6] <- "Total"
rownames(tt3)[6] <- "Total"
# move names into table
tt3 <- rbind(rownames(tt3), tt3)
tt3 <- cbind(rownames(tt3), tt3)
# name col/row by species
rownames(tt3) <- c("AcNC", "", "", "", "", "", "")
colnames(tt3) <- c("ApVT", "", "", "", "", "", "")

pandoc.table(tt3, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.", split.tables = "Inf")
```

Table 4 shows the number of transcripts that fall into each expression type for each each species. The totals for each species include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies. 



## Species comparison

In this section, I perform a number of comparisons of the thermal reactionomes between the *Ar* and *A22* species. Specifically, I:

- compare profiles of the temperature of maximum expression for thermally-responsive transcripts
- compare basal expression at optimal temperature for thermally-responsive genes
- compare thermal stability of *Intermediate* genes
- compare thermal sensitivity of *Bimodal* genes
- compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes
- evaluate the extent to which thermal reaction norms differ between species by mean and shape


### Compare temperature of maximum expression between species

Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*

```{r max_exp_PDF, warning = FALSE, message = FALSE}
# reshape data
Ap.df <- data.frame(Transcript = rep(Ap.response.type$Transcript, times = 2), colony = rep(c("ApVT", "AcNC"), each = length(Ap.response.type$Transcript)), max.val = c(Ap.response.type$A22.max, Ap.response.type$Ar.max))

maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
  geom_density(alpha=0.2, position="identity") + 
  labs(x = "Temperature of maximum expression", y = "Density") +
  theme(axis.title = element_text(size = rel(2))) +
  theme(axis.text = element_text(size = rel(1.2)))
print(maxexplot)

# Figure for presentation
png("results/temp_max_expression.png")
maxexplot + theme(legend.position = "none")
dev.off()
```


### Compare basal expression at optimal temperature for thermally-responsive genes

Genes upregulated in response to thermal stress in one species may have greater basal levels of expression in the other species that experiences those stressful conditions more often. To test this hypothesis, we compared expression levels near the optimal temperature (19.5C) between the two species for genes in that are either in the 'High' or 'Low' expression group in the other species. Specifically, do genes upregulated at high temperatures in *A22* have greater basal expression at optimal temperatures in *Ar*?

```{r optimum_expression_comparison}
# list of transcripts that are 'high' expressed in A22
A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]

# Compare expression at optimum temp (A22.opt) between colonies using t-test
t.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)

# T test on log-transformed values to control for outliers
t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
```

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`. As my goal is simply to determine if expression is typically greater at optimal temperatures (19.5 C) in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test

```{r A22_high_wilcoxon}
w1 <- wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
w1
```

Consistent with expectation, there is greater expression at 19.5C for *Ar* than *A22* transcripts for the set of transcripts that are transcripts that are up-regulated at high temperatures in *A22*. Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.

Next I test the converse: do genes up-regulated at low temperatures in *Ar* have greater basal expression near optimal temperatures in *A22*?

```{r Ar_low_wilcoxon}
# list of transcripts that are 'high' expressed in Ar
Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]

# t-test with log values
t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))

# Wilcoxon signed rank-test
w2 <- wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
w2
```

Counter to expectations, expression at optimal temperatures is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*. 

To confirm that there are not sample-level issues, I performed the same comparison using *Intermediate* expressed-genes where I do not expect to see a difference in expression.

```{r intermediate_transcripts}
# list of transcripts that are 'Intermediate' expressed in either colony
Ap.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" | Ap.response.type$A22.type == "Intermediate"), ]
# T test
t.test(log(Ap.int.transcripts$A22.opt+1), log(Ap.int.transcripts$Ar.opt+1))
# Wilcoxon signed rank-test
w3 <- wilcox.test(Ap.int.transcripts$A22.opt, Ap.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
w3
```

The non-parametric test for both comparisions also finds greater expression in *Ar* than *A22* at the optimal temperature.


### Compare thermal stability of *Intermediate* genes

*Intermediate* genes are those that have expression that is shut-down as conditions become stressful, likely non-essential molecular processes. We hypothesized that if the more southern *Ar* species was more thermally-tolerant than *A22*, transcripts with *Intermediate* expression (10-30C) would be active across a wider range of temperatures. To test this with our data, we calculated the standard deviation of the expression function for each temperature transcript that was 'Intermediate' expressed in each species.

```{r thermal_breadth}
# extract 'Intermediate' expressed transcripts for A22
A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
length(A22.int.lm)

# apply `transcriptSD` function to all transcripts
A22.int.sd <- unlist(Map(transcriptSD, A22.int.lm, colony = "A22"))
A22.int.sd <- data.frame(colony = "A22", exp_sd = A22.int.sd)

# repeat for Ar
Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
Ar.int.sd <- unlist(Map(transcriptSD, Ar.int.lm, colony = "Ar"))
Ar.int.sd <- data.frame(colony = "Ar", exp_sd = Ar.int.sd)

# T-test comparing standard deviation of expression between colonies
(t.varint <- t.test(Ar.int.sd$exp_sd, A22.int.sd$exp_sd, alternative = "two.sided"))
```

Consistent with our hypothesis, *Intermediate* transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*. 

```{r plot_thermal_breadth, echo=FALSE, eval=TRUE}
# prep data for plotting
Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)

# plot thermal breadth
g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g1 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression"))
```

### Compare thermal sensitivity of *Bimodal* genes

As the converse of the above hypothesis, a species that is especially thermally-sensitive is likely to activate expression of molecular processes for thermal tolerance more quickly. We tested this using the same approach as for the *Intermediate* transcripts, but using the inverse of the *Bimodal* expressed transcripts. 

```{r calc_Bimodal_variance}
# extract 'Bimodal' expressed transcripts for A22 species
A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
length(A22.bim.lm)

# apply `transcriptSD` function to all transcripts
A22.bim.sd <- unlist(Map(transcriptSD, A22.bim.lm, colony = "A22"))

# repeat for Ar 
Ar.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]]
length(Ar.bim.lm)
Ar.bim.sd <- unlist(Map(transcriptSD, Ar.bim.lm, colony = "Ar"))

# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
(t.varbim <- t.test(Ar.bim.sd, A22.bim.sd))
```

No difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.


### Compare the temperature at which gene expression increases, T~on~, for *High* and *Low* genes

Thermally-responsive genes could also differ in the temperatures at which they have increased or decreased expression in response to temperature changes. To examine this, I determine the temperature at which each responsive gene has the greated increase or decrease in expression.

```{r responsive_TPM}
# extract TPM data for thermally-responsive transcripts
resp.TPM.dt.sub <- TPM.dt.sub[names(responsive.lms)]
setkey(resp.TPM.dt.sub, Transcript)
str(resp.TPM.dt.sub)
length(unique(resp.TPM.dt.sub$Transcript))
# scale transcripts so can compare
resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
# rename colonies
resp.TPM.dt.sub$colony2 <- ifelse(resp.TPM.dt.sub$colony == "A22", "ApVT", "AcNC")
```

Predict expression for responsive transcripts.

```{r responsive_pred}
# apply predFunc to all responsive transcripts
resp.TPM.dt.sub.pred <- ddply(resp.TPM.dt.sub, .(Transcript), .inform="TRUE", predFunc)

# setkey to Transcript and colony 
resp.TPM.dt.sub.pred <- data.table(resp.TPM.dt.sub.pred)
setkey(resp.TPM.dt.sub.pred, Transcript, colony)
```

For next analyses, extract list of gene names by response type.

```{r extract_Transcripts}
A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), "Transcript"]
Ar.high.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "High"), "Transcript"]

A22.low.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Low"), "Transcript"]
Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), "Transcript"]

A22.bim.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]
Ar.bim.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]

A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), "Transcript"]
Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
```


Calculate T~on~ for *High* genes in each species.

```{r T_on_high}
# transcripts expressed at *High* and *Bimodal* temperatures in A22
A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.high.transcripts, A22.bim.transcripts), "A22")]
setkey(A22.high.TPM.dt.sub, Transcript)
str(A22.high.TPM.dt.sub)

# make data.frame for results
l1 <- length(unique(A22.high.TPM.dt.sub$Transcript))
A22.high.T_on <- data.frame(Transcript = unique(A22.high.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l1), type = rep("High", length = l1), T_on = NA)

# loop across transcripts, calculating T_on

for(i in unique(A22.high.TPM.dt.sub$Transcript)) {
    subdf <- A22.high.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val > 14), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    A22.high.T_on[which(A22.high.T_on$Transcript == i), "T_on"] <- T_on
}   

# repeat for Ar
Ar.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.high.transcripts, Ar.bim.transcripts), "Ar")]
setkey(Ar.high.TPM.dt.sub, Transcript)
str(Ar.high.TPM.dt.sub)

l2 <- length(unique(Ar.high.TPM.dt.sub$Transcript))
Ar.high.T_on <- data.frame(Transcript = unique(Ar.high.TPM.dt.sub$Transcript), colony = rep("AcNC", length = l2), type = rep("High", length = l2), T_on = NA)

for(i in unique(Ar.high.TPM.dt.sub$Transcript)) {
    subdf <- Ar.high.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val > 14), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "T_on"] <- T_on
}   

# determine if T~on~ is greater in *A22* or *Ar* for *High* genes.

(T_on.high.ttest <- t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on))
```

Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.

Repeat analysis for *Low* genes.

```{r T_on_low}
# transcripts expressed at *Low* temperatures in A22
A22.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(A22.low.transcripts, A22.bim.transcripts), "A22")]
setkey(A22.low.TPM.dt.sub, Transcript)
str(A22.low.TPM.dt.sub)

# make data.frame for results
l3 <- length(unique(A22.low.TPM.dt.sub$Transcript))
A22.low.T_on <- data.frame(Transcript = unique(A22.low.TPM.dt.sub$Transcript), colony = rep("ApVT", length = l3), type = rep("Low", length = l3), T_on = NA)

# loop across transcripts, calculating T_on
for(i in unique(A22.low.TPM.dt.sub$Transcript)) {
    subdf <- A22.low.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val < 21), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    A22.low.T_on[which(A22.low.T_on$Transcript == i), "T_on"] <- T_on
}   

# repeat for Ar
Ar.low.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(union(Ar.low.transcripts, Ar.bim.transcripts), "A22")]
setkey(Ar.low.TPM.dt.sub, Transcript)
str(Ar.low.TPM.dt.sub)

l4 <- length(unique(Ar.low.TPM.dt.sub$Transcript))
Ar.low.T_on <- data.frame(Transcript = unique(Ar.low.TPM.dt.sub$Transcript), colony = rep("AcNC", length = l4), type = rep("Low", length = length(unique(Ar.low.TPM.dt.sub$Transcript))), T_on = NA)

for(i in unique(Ar.low.TPM.dt.sub$Transcript)) {
    subdf <- Ar.low.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val < 21), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    Ar.low.T_on[which(Ar.low.T_on$Transcript == i), "T_on"] <- T_on
}   

(T_on.low.ttest <- t.test(Ar.low.T_on$T_on, A22.low.T_on$T_on))
```

Genes with increased expression at *Low* temperatures are on average turned on 0.2°C higher in *ApVT* than *AcNC*.

Visualize T~on~ for both *Low* and *High* genes on the same plot.

```{r plot_T_on, echo=FALSE, eval=TRUE}
# merge data for plotting
Ap.T_on <- rbind(A22.high.T_on, Ar.high.T_on, A22.low.T_on, Ar.low.T_on)
Ap.T_on$type <- factor(Ap.T_on$type, levels = c("Low", "High"))
Ap.T_on$Species <- ifelse(Ap.T_on$colony == "ApVT", "A. picea", "A. carolinensis")

# combined plot
T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = Species, group = Species, fill = Species)) +
    facet_grid(. ~ type) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(adjust = 3, fill = NA, size = 2) + 
    labs(x = "Temperature of gene activation", y = "Density") +
    theme(axis.title = element_text(size = rel(2))) +
    theme(axis.text = element_text(size = rel(1.2)))
#    theme(legend.position = "none")
print(T_on_plot)
```

### Evaluate the extent to which differences in thermal reaction norms are due to mean or shape

For the transcripts that differed in thermal responsiveness due to temperature, was the difference primarily due to differences in the mean value of expression, slope, curvature of a higher order effect? To test this, I rougly follow `r citet("10.1086/675302")` by defining differences among reation norms for individual genes due to changes in the overall mean, slope, curvature and all higher-order shape differences (i.e. wiggle).

- *Mean, M*: overall difference in the mean expression value across all temperatures
- *Slope, S*: difference in overall slope
- *Curvature, C*: average difference in curvature of the reaction norm
- *Wiggle, W*: variability in shape not captured by the previous three measures

```{r varshape, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.lms, .progress = "none", RxNvarshape)

# delta mean
boxplot(varshape.out$A22.mean, varshape.out$Ar.mean)
boxplot(log(varshape.out$A22.mean), log(varshape.out$Ar.mean))

t.mean <- t.test(log(varshape.out$A22.mean), log(varshape.out$Ar.mean), paired = TRUE)
t.mean

# delta slope
boxplot(log(varshape.out$A22.slope), log(varshape.out$Ar.slope))
t.slope <- t.test(log(varshape.out$A22.slope + 0.1), log(varshape.out$Ar.slope + 0.1), paired = TRUE)
t.slope

# delta curvature
boxplot(log(varshape.out$A22.curve), log(varshape.out$Ar.curve))
t.curvature <- t.test(log(varshape.out$A22.curve + 0.1), log(varshape.out$Ar.curve + 0.1), paired = TRUE)
t.curvature

# delta wiggle - small values so use un-transformed
boxplot(varshape.out$A22.wiggle, varshape.out$Ar.wiggle)
t.wiggle <- t.test(varshape.out$A22.wiggle, varshape.out$Ar.wiggle, paired = TRUE)
t.wiggle
```

Reaction norms differ between species by mean and slope, but not by curvature or wiggle.

Next, I partition the differences in the reaction norms into the variation explained by differences in the trait means and slope, as curvature and wiggle did not differ between species.

```{r partition_reaction_norm_variation}
# calculate differences in mean, slope, curvature and wiggle between colonies for each transcript
# mean = mean1 - mean2
varshape.out$lmean <- log(varshape.out$A22.mean) - log(varshape.out$Ar.mean)

# slope = slope1 - slope2
varshape.out$lslope <- log(varshape.out$A22.slope + 0.1) - log(varshape.out$Ar.slope + 0.1)

# take absolute value of each value and sum to get total differences between reaction norms
varshape.out$ltotal <- abs(varshape.out$lmean) + abs(varshape.out$lslope)

# variation in reaction norms due to differences in mean
varshape.out$prop.lmean <- abs(varshape.out$lmean) / varshape.out$ltotal
varshape.out$prop.lslope <- abs(varshape.out$lslope) / varshape.out$ltotal

# Mean proportion and 95% CI of total variation of each measure
mean(varshape.out$prop.lmean)
(qlmean <- quantile(varshape.out$prop.lmean, probs = c(0.05, 0.5, 0.95)))

mean(varshape.out$prop.lslope)
(qlslope <- quantile(varshape.out$prop.lslope, probs = c(0.05, 0.5, 0.95)))
```

From this analysis, about 80% of the differences in reaction norms between species are due to changes in the mean, with the remainder being due to changes in slope.



## Gene set enrichment analysis for thermal-responsive genes

### Functional annotation

In the previous section, I identified transcripts that show significant responses in expression. Next, I add gene annotation and ontology information to these transcripts.  

```{r annotation}
setkey(annotation.table, Sequence.Name)
signif.transcripts <- data.table(signif.transcripts)
setkey(signif.transcripts, Transcript)
signif.transcripts <- annotation.table[signif.transcripts]
setnames(signif.transcripts, "Sequence.Name", "Transcript")
```

```{r ann_responsive, echo=TRUE}
responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
# combine responsive.lms.ann with "type" information
responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
str(responsive.lms.ann.type)
write.csv(responsive.lms.ann.type, file = paste(resultsdir, "Ap_responsive_genes_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
```


### Proportion of responsive transcripts annotated

Of the responsive transcripts, `r prettyNum(length(which(responsive.lms.ann.type$best.hit.to.nr == "-"))/length(responsive.lms.ann.type$best.hit.to.nr)*100)`% are annotated, while `r prettyNum(length(which(annotation.table$best.hit.to.nr == "-"))/nrow(annotation.table) * 100)`% of all transcripts are annotated.


### Candidate gene enrichment

First, I explore of candidate genes with the GO term "response to stress" are enriched in the responsive data set. 

```{r GO_contingency_test}
# GO 'response to stress' hits in responsive transcripts
GO0006950.responsive <- responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]

# in high category
GO0006950.responsive[union(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]
# in low category
GO0006950.responsive[union(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
# in bimodal category
GO0006950.responsive[union(with(GO0006950.responsive, grep("Bimodal", Ar.type)), with(GO0006950.responsive, grep("Bimodal", A22.type))), ]
# in intermediate category
GO0006950.responsive[union(with(GO0006950.responsive, grep("Intermediate", Ar.type)), with(GO0006950.responsive, grep("Intermediate", A22.type))), ]

# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.lms compared to full list
resp.stress.responsive.count <- nrow(responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotation.table[grep("GO:0006950", annotation.table$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.lms.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotation.table) - resp.stress.all.count), nrow = 2)

GO.stress.Xsq <- chisq.test(GO.stress.table)
GO.stress.Xsq
```

While `r nrow(GO0006950.responsive)` "response to stress" genes are in the responsive set, this is not enriched compared to the whole transcriptome. 

```{r grep_shock}
hsp_responsive <- responsive.lms.ann.type[grep("shock", responsive.lms.ann.type$best.hit.to.nr), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
hsp_responsive

hsp_all <- annotation.table[grep("shock", annotation.table$best.hit.to.nr), list(Sequence.Name, best.hit.to.nr)]
nrow(hsp_all)
```

There are also `r nrow(hsp_responsive)` heat shock related genes in the responsive transcripts, out of `r nrow(hsp_all)` total heat shock related genes in the transcriptome annotation.


### Gene set enrichment analysis

I use [topGO](http://www.bioconductor.org/packages/2.12/bioc/html/topGO.html) to perform gene set enrichment analysis (GSEA) seperately for each expression type (bimodal, intermediate, high, low).

First need to create gene ID to GO term map file

```{r geneid2go, echo=TRUE, eval=TRUE, cache=TRUE}
# create geneid2go.map file from FastAnnotator annotationtable.txt
mapname = "results/geneid2go.map"
if(file.exists(mapname)) print("Gene ID to GO term file exists") else{
    geneid2GOmap(as.data.frame(annotation.table), ontology = c("BP", "CC", "MF"), mapname = mapname)
}
```

then read map file.

```{r readmap, echo=TRUE, eval=TRUE}
# read mappings file
geneID2GO <- readMappings(file = mapname)
str(head(geneID2GO))
```

### GSEA for thermally-responsive transcripts

Using this gene2GO map file, perform GSEA for:

**1) all responsive transcripts**

Use `selectFDR` function to select transcripts with adjusted P < 0.05.

Perform GSEA for all thermally-responsive transcripts.

For many of the responsive categories, it appears that there is considerable redundancy in the enriched GO terms. I use the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then perform hierarhichal clustering based on this information distance.

```{r gsea_all, results='hide'}
# create geneList. note that NA values cause problems with topGO
# so set any NA to 1 as need to retain for GO analysis
Ap.geneList <- RxNpval$padj
Ap.geneList[which(is.na(Ap.geneList))] <- 1
stopifnot(length(which(is.na(Ap.geneList))) == 0)
names(Ap.geneList) <- RxNpval$Transcript
str(Ap.geneList)

# Function to select genes with adjusted P-value < 0.05
selectFDR <- function(padj) {
    return(padj < 0.05)
}

# create topGOdata object
Ap.BP.GOdata <- new("topGOdata",
                 description = "BP gene set analysis", ontology = "BP",
                 allGenes = Ap.geneList, 
                 geneSel = selectFDR,
                 nodeSize = 10,
                 annot = annFUN.gene2GO, gene2GO = geneID2GO)

#Ap.BP.GOdata

# perform enrichment analysis using parentchild method
Ap.BP.resultParentChild <- runTest(Ap.BP.GOdata, statistic = 'fisher', algorithm = 'parentchild')
Ap.BP.resultParentChild

# table results
Ap.BP.ResTable <- GenTable(Ap.BP.GOdata, parentchild = Ap.BP.resultParentChild, topNodes = 10)
Ap.BP.ResTable
```

As the molecular processes involved in cold and hot temperature tolerance are different, I perform GSEA separately for each response type and compile results in a single table.

```{r gsea_Ap_high, results = 'hide'}
## Genes with *High* expression in both colonies
Ap.high <- Ap.response.type[which(Ap.response.type$Ar.type == "High" & Ap.response.type$A22.type == "High"), "Transcript"]
Ap.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.high) <- names(Ap.geneList)
Ap.geneList.high[(which(names(Ap.geneList.high) %in% Ap.high))] <- 1
# check correct number of values set to 1
table(Ap.geneList.high)
# run GSEA
Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)
```

```{r gsea_Ap_high_cluster}
# similarity among enriched GO terms
Ap.high.GO.term.sim <- termSim(Ap.high.gsea$GO.ID, Ap.high.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term
colnames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term

# distance matrix
Ap.high.hclust <- hclust(dist(Ap.high.GO.term.sim))
plot(Ap.high.hclust)

# report items in each cluster
Ap.high.hclust.report <- GSEAReportClusters(Ap.high.hclust, h = 1)

# add cluster information
Ap.high.gsea$Cluster <- NA

for(i in 1:length(Ap.high.hclust.report)) {
    Ap.high.gsea[Ap.high.hclust.report[[i]], "Cluster"] <- i
}
```

```{r gsea_Ap_low, results = "hide"}
## Genes with *Low* expression in both colonies
Ap.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Low" & Ap.response.type$A22.type == "Low"), "Transcript"]
Ap.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.low) <- names(Ap.geneList)
Ap.geneList.low[(which(names(Ap.geneList.low) %in% Ap.low))] <- 1
# check correct number of values set to 1
table(Ap.geneList.low)
# Run GSEA
Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
```

```{r gsea_Ap_low_cluster}
# similarity among enriched GO terms
Ap.low.GO.term.sim <- termSim(Ap.low.gsea$GO.ID, Ap.low.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ap.low.GO.term.sim) <- Ap.low.gsea$Term
colnames(Ap.low.GO.term.sim) <- Ap.low.gsea$Term

# distance matrix
Ap.low.hclust <- hclust(dist(Ap.low.GO.term.sim))
plot(Ap.low.hclust)

# report items in each cluster
Ap.low.gsea.cluster <- Ap.low.hclust.report <- GSEAReportClusters(Ap.low.hclust, h = 1.5)

# add cluster information
Ap.low.gsea$Cluster <- NA

for(i in 1:length(Ap.low.hclust.report)) {
    Ap.low.gsea[Ap.low.hclust.report[[i]], "Cluster"] <- i
}
```


```{r gsea_Ap_bim, resuts = 'hide'}
# Genes with *Bimodal* expression in both colonies
Ap.bim <- Ap.response.type[which(Ap.response.type$A22.type == "Bimodal" & Ap.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.bim) <- names(Ap.geneList)
Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
# check correct number of values set to 1
table(Ap.geneList.bim)
# Run GSEA
Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
```

```{r gsea_Ap_bim_cluster}
# similarity among enriched GO terms
Ap.bim.GO.term.sim <- termSim(Ap.bim.gsea$GO.ID, Ap.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ap.bim.GO.term.sim) <- Ap.bim.gsea$Term
colnames(Ap.bim.GO.term.sim) <- Ap.bim.gsea$Term

# distance matrix
Ap.bim.hclust <- hclust(dist(Ap.bim.GO.term.sim))
plot(Ap.bim.hclust)

# report items in each cluster
Ap.bim.hclust.report <- GSEAReportClusters(Ap.bim.hclust, h = 1.5)

# add cluster information
Ap.bim.gsea$Cluster <- NA

for(i in 1:length(Ap.bim.hclust.report)) {
    Ap.bim.gsea[Ap.bim.hclust.report[[i]], "Cluster"] <- i
}
```

```{r gsea_Ap_int, results = 'hide'}
# genes with *Intermediate* expression in both colonies
Ap.int <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate" & Ap.response.type$Ar.type == "Intermediate"), "Transcript"]
# create gene list, setting value to 1 for "int" transcripts
Ap.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.int) <- names(Ap.geneList)
Ap.geneList.int[(which(names(Ap.geneList.int) %in% Ap.int))] <- 1
# check correct number of values set to 1
table(Ap.geneList.int)
# run GSEA
Ap.int.gsea <- gsea(genelist = Ap.geneList.int, geneID2GO = geneID2GO)
```

```{r gsea_Ap_int_cluster}
# similarity among enriched GO terms
Ap.int.GO.term.sim <- termSim(Ap.int.gsea$GO.ID, Ap.int.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ap.int.GO.term.sim) <- Ap.int.gsea$Term
colnames(Ap.int.GO.term.sim) <- Ap.int.gsea$Term

# distance matrix
Ap.int.hclust <- hclust(dist(Ap.int.GO.term.sim))
plot(Ap.int.hclust)

# report items in each cluster
(Ap.int.hclust.report <- GSEAReportClusters(Ap.int.hclust, h = 1.5))

# add cluster information
Ap.int.gsea$Cluster <- NA

for(i in 1:length(Ap.int.hclust.report)) {
    Ap.int.gsea[Ap.int.hclust.report[[i]], "Cluster"] <- i
}
```

```{r Ap_gsea_intersect_save, echo = FALSE}
Ap.high.gsea$Type <- "High"
Ap.low.gsea$Type <- "Low"
Ap.bim.gsea$Type <- "Bimodal"
Ap.int.gsea$Type <- "Intermediate"
Ap.gsea.intersect <- rbind(Ap.low.gsea, Ap.int.gsea, Ap.high.gsea, Ap.bim.gsea)

Ap.gsea.intersect <- Ap.gsea.intersect[,c("Type", "GO.ID", "Term", "Cluster", "Annotated", "Significant", "Expected", "parentchild")]
colnames(Ap.gsea.intersect)[which(colnames(Ap.gsea.intersect) == "parentchild")] <- "P"

write.csv(Ap.gsea.intersect, file = paste(resultsdir, "Ap_gsea_intersect_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
```

GSEA for genes that are *Intermediate* in A. carolinensis and *Bimodal* in A. picea

```{r gsea_Ar_int_A22_bim, results = 'hide'}
# genes with *Intermediate* expression in Ar and *Bimodal* in A22
Ar.int.A22.bim <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Bimodal"), "Transcript"]
Ar.int.A22.bim.geneList <- rep(0, length = length(Ap.geneList))
names(Ar.int.A22.bim.geneList) <- names(Ap.geneList)
Ar.int.A22.bim.geneList[(which(names(Ar.int.A22.bim.geneList) %in% Ar.int.A22.bim))] <- 1
# check correct number of values set to 1
table(Ar.int.A22.bim.geneList)
# run GSEA
Ar.int.A22.bim.gsea <- gsea(genelist = Ar.int.A22.bim.geneList, geneID2GO = geneID2GO, plotpath = NA)
```

Cluster by semantic similarity

```{r gsea_Ar_int_A22_bim_cluster}
# similarity among enriched GO terms
Ar.int.A22.bim.gsea.term.sim <- termSim(Ar.int.A22.bim.gsea$GO.ID, Ar.int.A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
colnames(Ar.int.A22.bim.gsea.term.sim) <- Ar.int.A22.bim.gsea$Term
# distance matrix
Ar.int.A22.bim.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.bim.gsea.term.sim))
plot(Ar.int.A22.bim.gsea.term.sim.hclust)

# report items in each cluster at height 1.5
(Ar.int.A22.bim.hclust.report <- GSEAReportClusters(Ar.int.A22.bim.gsea.term.sim.hclust, h = 1))
```

GSEA for genes that are *Intermediate* in A. carolinensis and *Low* in A. picea

```{r gsea_Ar_int_A22_low, results = 'hide'}
# genes with *Intermediate* expression in Ar and *Low* in A22
Ar.int.A22.low <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate" & Ap.response.type$A22.type == "Low"), "Transcript"]
Ar.int.A22.low.geneList <- rep(0, length = length(Ap.geneList))
names(Ar.int.A22.low.geneList) <- names(Ap.geneList)
Ar.int.A22.low.geneList[(which(names(Ar.int.A22.low.geneList) %in% Ar.int.A22.low))] <- 1
# check correct number of values set to 1
table(Ar.int.A22.low.geneList)
# run GSEA
Ar.int.A22.low.gsea <- gsea(genelist = Ar.int.A22.low.geneList, geneID2GO = geneID2GO, plotpath = NA)
```

```{r gsea_Ar_int_A22_low_cluster}
# similarity among enriched GO terms
Ar.int.A22.low.gsea.term.sim <- termSim(Ar.int.A22.low.gsea$GO.ID, Ar.int.A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
colnames(Ar.int.A22.low.gsea.term.sim) <- Ar.int.A22.low.gsea$Term
# distance matrix
Ar.int.A22.low.gsea.term.sim.hclust <- hclust(dist(Ar.int.A22.low.gsea.term.sim))
plot(Ar.int.A22.low.gsea.term.sim.hclust)
# report items in each cluster at height 1
(Ar.int.A22.low.hclust.report <- GSEAReportClusters(Ar.int.A22.low.gsea.term.sim.hclust, h = 1.2))
```




### GSEA for each functional type in each species

GO enrichment for *A. picea* **High** genes.

```{r gsea_A22_high, results = 'hide'}
# A22 'High' genes
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high.transcripts))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
```
Clustering of enriched GO terms for *A. picea* **High** genes.

```{r gsea_A22_high_cluster}
# similarity among enriched GO terms
A22.high.gsea.term.sim <- termSim(A22.high.gsea$GO.ID, A22.high.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
colnames(A22.high.gsea.term.sim) <- A22.high.gsea$Term
# distance matrix
A22.high.gsea.term.sim.hclust <- hclust(dist((A22.high.gsea.term.sim)))
plot((A22.high.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.high.gsea.term.sim.hclust, h = 1.2)
```

GO enrichment for *A. picea* **Low** genes.

```{r gsea_A22_low, results = 'hide'}
# A22 'Low' genes
A22.geneList.low <- rep(0, length = length(Ap.geneList))
names(A22.geneList.low) <- names(Ap.geneList)
A22.geneList.low[(which(names(A22.geneList.low) %in% A22.low.transcripts))] <- 1
A22.low.gsea <- gsea(genelist = A22.geneList.low, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. picea* **Low** genes.

```{r gsea_A22_low_cluster}
# similarity among enriched GO terms
A22.low.gsea.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
colnames(A22.low.gsea.term.sim) <- A22.low.gsea$Term
# distance matrix
A22.low.gsea.term.sim.hclust <- hclust(dist((A22.low.gsea.term.sim)))
plot((A22.low.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.low.gsea.term.sim.hclust, h = 2)
```

GO enrichment for *A. picea* **Bimodal** genes.

```{r gsea_A22_bim, results = 'hide'}
# A22 'Bim' genes
A22.geneList.bim <- rep(0, length = length(Ap.geneList))
names(A22.geneList.bim) <- names(Ap.geneList)
A22.geneList.bim[(which(names(A22.geneList.bim) %in% A22.bim.transcripts))] <- 1
A22.bim.gsea <- gsea(genelist = A22.geneList.bim, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. picea* **Bimodal** genes.

```{r gsea_A22_bim_cluster}
# similarity among enriched GO terms
A22.bim.gsea.term.sim <- termSim(A22.bim.gsea$GO.ID, A22.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
colnames(A22.bim.gsea.term.sim) <- A22.bim.gsea$Term
# distance matrix
A22.bim.gsea.term.sim.hclust <- hclust(dist((A22.bim.gsea.term.sim)))
plot((A22.bim.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.bim.gsea.term.sim.hclust, h = 1)
```

GO enrichment for *A. picea* **Intermediate** genes.

```{r gsea_A22_int, results = 'hide'}
# A22 'Int' genes
A22.geneList.int <- rep(0, length = length(Ap.geneList))
names(A22.geneList.int) <- names(Ap.geneList)
A22.geneList.int[(which(names(A22.geneList.int) %in% A22.int.transcripts))] <- 1
A22.int.gsea <- gsea(genelist = A22.geneList.int, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. picea* **Intermediate** genes.

```{r gsea_A22_int_cluster}
# similarity among enriched GO terms
A22.int.gsea.term.sim <- termSim(A22.int.gsea$GO.ID, A22.int.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
colnames(A22.int.gsea.term.sim) <- A22.int.gsea$Term
# distance matrix
A22.int.gsea.term.sim.hclust <- hclust(dist((A22.int.gsea.term.sim)))
plot((A22.int.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(A22.int.gsea.term.sim.hclust, h = 1.3)
```

GO enrichment for *A. carolinensis* **High** genes.

```{r gsea_Ar_high, results = 'hide'}
# Ar 'High' genes
Ar.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.high) <- names(Ap.geneList)
Ar.geneList.high[(which(names(Ar.geneList.high) %in% Ar.high.transcripts))] <- 1
Ar.high.gsea <- gsea(genelist = Ar.geneList.high, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. carolinensis* **High** genes.

```{r gsea_Ar_high_cluster}
# similarity among enriched GO terms
Ar.high.gsea.term.sim <- termSim(Ar.high.gsea$GO.ID, Ar.high.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.high.gsea.term.sim) <- Ar.high.gsea$Term
colnames(Ar.high.gsea.term.sim) <- Ar.high.gsea$Term
# distance matrix
Ar.high.gsea.term.sim.hclust <- hclust(dist((Ar.high.gsea.term.sim)))
plot((Ar.high.gsea.term.sim.hclust))
# report items in each cluster
GSEAReportClusters(Ar.high.gsea.term.sim.hclust, h = 1.2)
```
GO enrichment for *A. carolinensis* **Low** genes.

```{r gsea_Ar_low, results = 'hide'}
# Ar 'Low' genes
Ar.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.low) <- names(Ap.geneList)
Ar.geneList.low[(which(names(Ar.geneList.low) %in% Ar.low.transcripts))] <- 1
Ar.low.gsea <- gsea(genelist = Ar.geneList.low, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. carolinensis* **Low** genes.

```{r gsea_Ar_low_cluster}
# similarity among enriched GO terms
Ar.low.gsea.term.sim <- termSim(Ar.low.gsea$GO.ID, Ar.low.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.low.gsea.term.sim) <- Ar.low.gsea$Term
colnames(Ar.low.gsea.term.sim) <- Ar.low.gsea$Term
# distance matrix
Ar.low.gsea.term.sim.hclust <- hclust(dist((Ar.low.gsea.term.sim)))
plot((Ar.low.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(Ar.low.gsea.term.sim.hclust, h = 2)
```

GO enrichment for *A. carolinensis* **Bimodal** genes.

```{r gsea_Ar_bim, results = 'hide'}
# Ar 'Bim' genes
Ar.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.bim) <- names(Ap.geneList)
Ar.geneList.bim[(which(names(Ar.geneList.bim) %in% Ar.bim.transcripts))] <- 1
Ar.bim.gsea <- gsea(genelist = Ar.geneList.bim, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. carolinensis* **Bimodal** genes.

```{r gsea_Ar_bim_cluster}
# similarity among enriched GO terms
Ar.bim.gsea.term.sim <- termSim(Ar.bim.gsea$GO.ID, Ar.bim.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.bim.gsea.term.sim) <- Ar.bim.gsea$Term
colnames(Ar.bim.gsea.term.sim) <- Ar.bim.gsea$Term
# distance matrix
Ar.bim.gsea.term.sim.hclust <- hclust(dist((Ar.bim.gsea.term.sim)))
plot((Ar.bim.gsea.term.sim.hclust))
# report items in each cluster at height 1.8
GSEAReportClusters(Ar.bim.gsea.term.sim.hclust, h = 1)
```

GO enrichment for *A. carolinensis* **Intermediate** genes.

```{r gsea_Ar_int, results = 'hide'}
# Ar 'Int' genes
Ar.geneList.int <- rep(0, length = length(Ap.geneList))
names(Ar.geneList.int) <- names(Ap.geneList)
Ar.geneList.int[(which(names(Ar.geneList.int) %in% Ar.int.transcripts))] <- 1
Ar.int.gsea <- gsea(genelist = Ar.geneList.int, geneID2GO = geneID2GO)
```

Clustering of enriched GO terms for *A. carolinensis* **Intermediate** genes.

```{r gsea_Ar_int_cluster}
# similarity among enriched GO terms
Ar.int.gsea.term.sim <- termSim(Ar.int.gsea$GO.ID, Ar.int.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ar.int.gsea.term.sim) <- Ar.int.gsea$Term
colnames(Ar.int.gsea.term.sim) <- Ar.int.gsea$Term
# distance matrix
Ar.int.gsea.term.sim.hclust <- hclust(dist((Ar.int.gsea.term.sim)))
plot((Ar.int.gsea.term.sim.hclust))
# report items in each cluster
GSEAReportClusters(Ar.int.gsea.term.sim.hclust, h = 1.8)
```

Combine gene set enrichment analysis for each species and category into single table.

```{r gsea_combine, eval = TRUE}
# combine into single table
A22.high.gsea$Type <- "High"
A22.low.gsea$Type <- "Low"
A22.int.gsea$Type <- "Intermediate"
A22.bim.gsea$Type <- "Bimodal"
A22.gsea <- rbind(A22.high.gsea, A22.low.gsea, A22.int.gsea, A22.bim.gsea)
A22.gsea$Species <- "ApVT"

Ar.high.gsea$Type <- "High"
Ar.low.gsea$Type <- "Low"
Ar.bim.gsea$Type <- "Bimodal"
Ar.int.gsea$Type <- "Intermediate"
Ar.gsea <- rbind(Ar.high.gsea, Ar.low.gsea, Ar.int.gsea, Ar.bim.gsea)
Ar.gsea$Species <- "AcNC"

# combine
Ap.gsea.union <- rbind(A22.gsea, Ar.gsea)
# reorder
Ap.gsea.union <- Ap.gsea.union[,c("Species", "Type", "GO.ID", "Term", "Annotated", "Significant", "Expected", "parentchild")]
colnames(Ap.gsea.union)[8] <- "P"

write.csv(Ap.gsea.union, file = paste(resultsdir, "Ap_gsea_union", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
```



## Visualize responsive transcripts

Plots for all genes expressed at *High* temps in GO category "GO:0006950: response to stress"

```{r plot_GOstress, echo=FALSE, eval=TRUE}
high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]

# in low category
low.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]
# in bimodal category
bim.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Bimodal", Ar.type)), with(GO0006950.responsive, grep("Bimodal", A22.type))), ]
# in intermediate category
int.genes <- GO0006950.responsive[union(with(GO0006950.responsive, grep("Intermediate", Ar.type)), with(GO0006950.responsive, grep("Intermediate", A22.type))), ]


# Line plot, expression against temp, faceted by species
trp1 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_line() + 
  facet_grid(. ~ colony2) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp1)

# Smooth plot, expression against temp, faceted by colony
trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony2) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp2)
```

Make plots for all genes expressed at *Low* temps in GO category "GO:0006950: response to stress"

```{r plot_GOstress_low, echo=FALSE, eval=TRUE, cache=TRUE}
low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]

# Line plot, expression against temp, faceted by colony
trp3 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_line() + 
  facet_grid(. ~ colony2) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp3)

# Smooth plot, expression against temp, faceted by colony
trp4 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony2) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp4)
```


```{r ggplot_high}
trp_A22_high <- ggplot(resp.TPM.dt.sub[A22.high.transcripts][1:220,], aes(x=val, y=TPM.scaled, group=Transcript)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony2) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp_A22_high)

### predicted values
resp.TPM.dt.sub.pred[,pTPM.scaled:=scale(pTPM), by = Transcript]

pred_A22_high <- ggplot(resp.TPM.dt.sub.pred[A22.high.transcripts][1:220,], aes(x=val, y=pTPM.scaled, group=Transcript)) +
  geom_line() + 
  facet_grid(. ~ colony) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(pred_A22_high)

png("RxN_example.png")
ggplot(resp.TPM.dt.sub.pred["100348|*|comp3450455_c0_seq1"], aes(x=val, y=pTPM.scaled, group=Transcript, color = Transcript)) +
  geom_line() + 
  facet_grid(. ~ colony) +
  labs(y = "Expression (scaled)", x = expression("Temperature ", degree, "C")) +
  theme(legend.position = "none")
dev.off()
```

## Shiny interactive web-app

To assist visualization of specific transcripts, I made a interactive web-app using the [shiny](http://www.rstudio.com/shiny/) package. The scripts for this app are in the sub-directory `.\ApRxN-shinyapp`.

Export data for interactive shiny app. 

```{r shiny_file, eval=FALSE, echo=FALSE}
shiny.dt <- TPM.dt.sub[responsive.lms.ann.type[,list(Transcript, best.hit.to.nr.1, GO.Biological.Process.1)]]

# scale expression values 
shiny.dt[,TPM.scaled:=scale(TPM), by = Transcript]
str(shiny.dt)
write.csv(shiny.dt, file = "ApRxN-shinyapp/responsive.lms.TPM.csv", quote = TRUE, row.names = FALSE)
```


```{r rData, echo=FALSE}
# save RData for manuscript file
save(TPM.dt.sub, RxNpval, sigtable, chi1, resp.TPM.dt.sub, Ap.type.table, tt2, tt3, qlslope, qlmean, GO.stress.Xsq, GO0006950.responsive, Ap.df, w1, w2, w3, Ap.T_on, Ap.BP.resultParentChild, resp.stress.responsive.count, t.mean, t.slope, t.curvature, t.wiggle, t.varint, t.varbim, T_on.high.ttest, T_on.low.ttest, Ap.gsea.union, Ap.gsea.intersect, Ar.int.A22.bim.gsea, Ar.int.A22.low.gsea, file = "ms.RData")
```


## Session information

```{r session}
save.image()
sessionInfo()
```

## References

```{r references, echo = FALSE, results='asis'}
bibliography()
```
